The GitHub Open Source Development Process \author{Kevin Peterson\\ Department of Information Technology\\ Clinical Informatics Support Systems\\ Mayo Clinic\\ 200 1st Street\\ Rochester, MN 55905\\ \small \texttt{peterson.kevin@mayo.edu} }

### 2013-04-18 paper/main.tex
Added the GitHub Staff interview transcript (Thanks to John Britton!).
Finished adding in Rick's comments.

By analyzing a subset of GitHub repositories, this report will show how GitHub has influenced some intrinsic aspects of traditional OSS development, such as developer hierarchies and issue close velocity. We find that many of the traditional aspects of OSS development remain, such as most project development being done by a small group of core developers. Other traditional assumptions about OSS project hierarchy, such as a large number of Issue Reporters compared to Committers, seems unsupported by the GitHub data. We conclude that GitHub represents an evolution of the OSS development process, and not necessarily a large shift.
\end{abstract}

@@

In addition, three supporting research questions are proposed. These questions will draw upon data from this work, but also reference several case studies to better represent how GitHub may be changing the way in which OSS development is conducted. These qualitative observations, paired with the more qualitative data analysis, allows for a broad description of the way that GitHub is changing the Software Engineering landscape for OSS.

@@

The Apache Project noted that a rapid response to problems can be obtained because OSS is not bound to release schedules in the same way commercial software is. ``Patches'' may be released at any time, by any member of the community\cite{mockus2000case}. ``Patches'' in the context of GitHub could be equated to \emph{Pull Requests}. A \emph{Pull Request} is an announcement by a \emph{fork} that there have been changes made that may be appropriate to add back to the main repository. The owners of the main repository may then audit these requests, and \emph{pull} some or all of them back into the main repository.

@@

\noindent \textbf{Research Question 2: How is GitHub changing the OSS process?}\\
The social aspects of GitHub are an important part of the development experience\cite{dabbish2012social}. User interactions and social pressures can drive OSS development in interesting ways. Social aspects of OSS development have existed before GitHub, with mailing lists\cite{mockus2000case}, gift giving mechanisms\cite{bergquist2008power}, and other code hosting sites. What is GitHub doing to facilitate change on OSS development, and how does it see OSS development changing moving forward?\\

\noindent \textbf{Research Question 3: Can GitHub be used for more that code artifacts?}\\
The Object Management Group\textsuperscript{\textregistered}\footnote{http://www.omg.org} is a computer industry consortium focused on technology standards. Potential standard submission teams must go through a vetting process\cite{kobryn1999uml} which allows industry representatives to provide feedback on the standard. Gathering and recording this feedback is a challenge, but GitHub's issue tracking system may be able to streamline the process. This just one example of how GitHub is being used for non-code artifacts. Are organizations exploring this type of work flow?

\section{Methods}
\subsection{Collection and Storage}
Data collection was done using the GitHub REST API.\footnote{http://developer.github.com/v3/} A random selection of 1000 repositories was selected, and metrics were gathered. The selection process used for choosing the repository sample is described in Algorithm \ref{algo:data_collection_algo}.

@@

Repositories were gathered at random, but only repositories owned by an \textit{Organization} GitHub User type were considered. After a random repository has been selected, the \textsc{isOrgRepo} function will filter out all non \textit{Organization} owned repositories. Next, the \textsc{store} function accepts as an input the given repository. The purpose of this function is to persist the given repository to a database for further analysis. Results were stored in a MySQL relational database management system. The database schema to store the data is described in Figure \ref{fig:er_diagram}.

@@

\noindent \textbf{\textit{Commits:}}
The number of individual source code commits to a repository. It is important to note that a \textit{commit} is a single transaction that changes the code contents of a repository. Actual commit ``size'' (lines of code modified, for example) is not taken into account.

@@

The data reinforces the hypothesis, as does other research in the area. Even though there are arguments that the number of Forks is the true measure of project success\cite{baudry2012towards}, the data suggests that the two measures are related. Also, \textit{Dabbish et al.} have noted that the number of Watches can be a key for others to gauge interest in a project, and thus determine if they should participate \cite{dabbish2013leveraging}. If the number of Watchers of a repository serves as a social cue for participation, even ``passive'' involvement with a project (Watching) could influence the ``active'' participation (Forking). This could possibly contribute to the correlation of the two variables.\\

@@

Figure \ref{fig:issue_reporters_committers_scatterplot}, does however suggest, however, that that the two variables may be related. The hypothesis implies that there will be a correlation between the two variables -- which is demonstrated by a reasonably strong correlation coefficient of \input{issue_reporters_committers_scatterplot-correlation}}\unskip. More investigation would be necessary to further examine this finding.\\

\subsection{Research Questions}
\noindent \textbf{Research Question 1: Are GitHub projects primarily focused around a small set of core Committers?}\\
The data in Figure \ref{fig:committers_percentage_pie_chart} represents commit percentage breakdown. The intent of this measurement is to show how the number of repository commits are distributed among it's committers. The slices represent the distribution of committers that that have committed the outer percentage amount of commits to a project, or:

@@

Despite the potential migration cost, the National Center for Biomedical Ontology,\footnote{https://github.com/ncbo} the National Cancer Informatics Program,\footnote{https://github.com/ncip} NASA,\footnote{https://github.com/nasa} and the WhiteHouse\footnote{https://github.com/WhiteHouse} are just some organizations that have moved all or portions of development to GitHub. As these and other organizations consider the move to GitHub, the distributed and ``social'' aspects of the OSS process will continue to evolve.\\

\noindent \textbf{Research Question 3: Can GitHub be used for more that code artifacts?}\\
The OMG\textsuperscript{\textregistered} standardization process is a process by which products strive to produce an interoperability standard by way of an industry standards consortium\cite{omgTechProcess}.
The standardization process itself aims to, at its conclusion, produce a set of model artifacts containing sufficient detail such that when interpreted and implemented, produce interoperable software.
At a conceptual level, this can be thought of as a software requirements engineering process as much as a standardization process. If so, of the five main tasks of requirements engineering, Elicitation, Analysis and Negotiation, Documentation, Validation, and Management,\cite{sommerville1998requirements} we can show that GitHub is an appropriate forum for many of them.

@@

\section{Conclusion}
The data suggests an evolution in process, but not necessarily a large shift. A small subset of core developers, a traditional aspect of OSS development\cite{mockus2000case,mockus2002two,krishnamurthy2002cave}, seems supported by both the summary data in Figure \ref{eq:variance}, and the distribution of committers per repository in Figure \ref{fig:committers_histogram}. Figure \ref{fig:committers_percentage_pie_chart} shows that a significant number of participation, albeit small, in many repositories. This may indicate a lowered entry barrier for participation.

The data summary variance statement in Assertion: \ref{eq:variance} is important, as it suggests large amounts of variety in the observed variables. This may suggest that not all repositories on GitHub are intended to be consumable software artifacts for users, or ``consumers.'' A large number of repositories may not be intended for customer consumption, but merely as example code, experimentation, or exercises. More research into categorizing GitHub repositories is needed. This is one reason why only repositories owned by a GitHub \textit{Organization} were analyzed, as it was anticipated that this would yield a better sample of repositories for analysis.

The non-code artifacts in GitHub may fall outside of the domain of software development, but it is a significant aspect of our findings. From government documents to software specifications, GitHub is being used and used successfully. More research is needed into what role GitHub will play in the work flow of these artifacts, and whether it will influence their traditional work flow, as it has for software development.

\section{Acknowledgments}
We thank Gilberto Fragoso, Ph.D., Paul Alexander, and the GitHub Staff for insight and ideas, Harold Solbrig, Craig Stancl, and Dick Hedger for guidance, and Rick Kiefer for his review.

@@

%\input{interview}

>>>>

By analyzing a subset of GitHub repositories, this report will show how GitHub has influenced some intrinsic aspects of traditional OSS development, such as developer hierarchies and issue close velocity. We find that many of the traditional aspects of OSS development remain, such as most project development being done by a small group of core developers. Other traditional assumptions about OSS developer hierarchy, such as a large number of Issue Reporters compared to Committers, seems unsupported by the GitHub data. We conclude that GitHub represents an evolution of the OSS development process, and not necessarily a large shift.
\end{abstract}

@@

In addition, three supporting research questions are proposed. These questions will draw upon data from this work, but also reference several case studies to better represent how GitHub may be changing the way in which OSS development is conducted. These qualitative observations, paired with the more quantitative data analysis, allows for a broad description of the way that GitHub is changing the Software Engineering landscape for OSS.

@@

The Apache Project noted that a rapid response to problems can be obtained because OSS is not bound to release schedules in the same way as commercial software. ``Patches'' may be released at any time, by any member of the community\cite{mockus2000case}. ``Patches'' in the context of GitHub could be equated to \emph{Pull Requests}. A \emph{Pull Request} is an announcement by a \emph{fork} that there have been changes made that may be appropriate to add back to the main repository. The owners of the main repository may then audit these requests, and \emph{pull} some or all of them back into the main repository.

@@

\noindent \textbf{Research Question 2: How is GitHub changing the OSS process?}\\
The social aspects of GitHub are an important part of the development experience\cite{dabbish2012social}. User interactions and social pressures can drive OSS development in interesting ways. Social aspects of OSS development have existed before GitHub, with mailing lists\cite{mockus2000case}, gift giving mechanisms\cite{bergquist2008power}, and other collaboration tools. What is GitHub doing to facilitate change in OSS development, and how does it see OSS development changing moving forward?\\

\noindent \textbf{Research Question 3: Can GitHub be used for more than code artifacts?}\\
The Object Management Group\textsuperscript{\textregistered}\footnote{http://www.omg.org} is a computer industry consortium focused on technology standards. Potential standard submission teams must go through a vetting process\cite{kobryn1999uml} which allows industry representatives to provide feedback on the standard. Gathering and recording this feedback is a challenge, but GitHub's issue tracking system may be able to streamline the process. This just one example of how GitHub is being used for non-code artifacts. Are organizations exploring this type of work flow?

\section{Methods}
\subsection{Collection and Storage}
Data collection was done using the GitHub REST API.\footnote{http://developer.github.com/v3/} A random selection of 1000 repositories was selected, and metrics were gathered. Only repositories owned by a GitHub \textit{Organization} were considered. GitHub \textit{Organizations} are group-owned accounts\footnote{https://github.com/blog/674-introducing-organizations}, as opposed to normal individual user-owned accounts. \textit{Organization} owned repositories were chosen to increase the likelihood of analyzing high-participation projects. The selection process used for choosing the repository sample is described in Algorithm \ref{algo:data_collection_algo}.

@@

When a random repository has been selected, the \textsc{isOrgRepo} function will filter out all non \textit{Organization} owned repositories. Next, the \textsc{store} function accepts as an input the given repository. The purpose of this function is to persist the given repository to a database for further analysis. Results were stored in a MySQL relational database management system. The database schema to store the data is described in Figure \ref{fig:er_diagram}.

@@

\noindent \textbf{\textit{Commits:}}
The number of individual source code commits to a repository. It is important to note that a \textit{commit} is a single transaction that changes the code contents of a repository. Actual commit ``size'' (lines of code modified, number of files changed, for example) is not taken into account.

@@

The data reinforces the hypothesis, as does other research in the area. Even though there are arguments that the number of Forks is the true measure of project success\cite{baudry2012towards}, the data suggests that the two measures are related. Also, \textit{Dabbish et al.} have noted that the number of Watches can be a key for others to gauge interest in a project, and thus determine if they should participate \cite{dabbish2013leveraging}. If the number of Watchers of a repository serves as a social cue for participation, even ``passive'' involvement with a project (Watching) could influence the ``active'' participation (Forking). This could possibly contribute to the correlation of the two variables. It is important to note, however, that we cannot conclude any causal relationship between the two variables. Concluding whether the social cue for participation is the Forks, the Watchers, or some combination, is not possible given the data and methods used.\\

@@

Figure \ref{fig:issue_reporters_committers_scatterplot}, does however suggest, however, that that the two variables are related. The hypothesis implies that there will be a correlation between the two variables -- which is demonstrated by a reasonably strong correlation coefficient of \input{issue_reporters_committers_scatterplot-correlation}}\unskip. More investigation would be necessary to further examine this finding.\\

\subsection{Research Questions}
\noindent \textbf{Research Question 1: Are GitHub projects primarily focused around a small set of core Committers?}\\
The data in Figure \ref{fig:committers_percentage_pie_chart} represents commit percentage breakdown, or an analysis of developer \textit{Contribution} percentages to a repository. The intent of this measurement is to show how the number of repository commits are distributed among it's committers. The slices represent the distribution of committers that have committed a given percentage of commits (or a \textit{Contribution} percentage, as labeled in the figure) to a project. For example, a small percentage of developers have contributed $95-100\%$ of the commits to a repository, so this is represented as a small slice in the figure. Contrast that to developers that have committed $0-5\%$ of the commits to a repository. This type of contribution is common, and represented as a larger slice. The \textit{Contribution} percentages are calculated by:

@@

Despite the potential migration cost, the National Center for Biomedical Ontology,\footnote{https://github.com/ncbo} the National Cancer Informatics Program,\footnote{https://github.com/ncip} NASA,\footnote{https://github.com/nasa} and the WhiteHouse\footnote{https://github.com/WhiteHouse} are just some organizations that have moved all or portions of development to GitHub. As these and other organizations consider the move to GitHub, the distributed and ``social'' aspects of the OSS process will continue to evolve.

According to GitHub Staff member John Britton, lowering the barrier to participation is an important cultural shift:
\begin{quote}
GitHub moved the open source community away from a permission culture. Prior to GitHub, contributing to an open source project was very difficult and often involved asking permission. By allowing people to fork a repository and make changes without asking for permission, GitHub has made contributing to someone else's project much easier than it has ever been. By submitting a Pull Request with suggested changes, project maintainers and contributors have a place to discuss changes to code that is already written and to refine it.\cite{github_interview}
\end{quote}

Spending large amounts of time and resources to participate is a detriment to knowledge-sharing in general, not just OSS development\cite{ardichvili2003motivation}. Because of this, we conclude that this permissive culture of knowlege-sharing that GitHub fosters represents an important change in the OSS process.

\noindent \textbf{Research Question 3: Can GitHub be used for more than code artifacts?}\\
The OMG\textsuperscript{\textregistered} standardization process is a process by which products strive to produce an interoperability standard by way of an industry standards consortium\cite{omgTechProcess}.
The standardization process itself aims to, at its conclusion, produce a set of model artifacts containing sufficient detail such that when interpreted and implemented, produce interoperable software.
At a conceptual level, this can be thought of as a software requirements engineering process as much as a standardization process. If so, of the five main tasks of requirements engineering, Elicitation, Analysis and Negotiation, Documentation, Validation, and Management,\cite{sommerville1998requirements} we can show that GitHub is an appropriate forum for many of them.

@@

\section{Conclusion}
The data suggests an \textit{evolution} in process, but not necessarily a large shift. A small subset of core developers, a traditional aspect of OSS development\cite{mockus2000case,mockus2002two,krishnamurthy2002cave}, seems supported by both the summary data in Figure \ref{eq:variance}, and the distribution of committers per repository in Figure \ref{fig:committers_histogram}. Figure \ref{fig:committers_percentage_pie_chart} shows a significant number of participators, albeit with small contribution amounts, in many repositories. This may indicate a lowered entry barrier for participation, and is consistent with the conclusions reached in Research Question 2.

The data does not reflect some traditional aspects of OSS development. Notably, it was not observed that there were far more \emph{Issue Reporters} than code \emph{Committers}, which is contrary to what the Apache Server report concluded\cite{mockus2000case}. A possible explanation is the amount of variability in the data collected. Assertion \ref{eq:variance}, which is derived from the summary data shown in Table \ref{table:summary_stats}, suggests large amounts of variety in the observed variables. This may suggest a wide range of repository purposes, or specifically, that not all repositories on GitHub are intended to be consumable software artifacts for users, or ``customers.'' A large number of repositories may not be intended for customer consumption, but merely as example code, experimentation, or exercises. Given this, we cannot state any conclusions regarding participation rates of the \emph{Issue Reporters}. More research into categorizing GitHub repositories, and the role of \emph{Issue Reporters} in them, is needed.

The most notable \textit{evolution} in OSS process driven by GitHub, we conclude, is the lowered barrier to participation. Qualitative analysis conducted in Research Question 2 suggest a cultural shift\cite{begel2013social,github_interview} in the participation process. The data also supports high levels of participation, especially Figure \ref{fig:committers_percentage_pie_chart}, which shows that many repositories have large numbers of developers committing small amounts of code. This, we conclude, represents an important \textit{evolution} in OSS development process.

\section{Acknowledgments}
We thank Gilberto Fragoso, Ph.D., Paul Alexander, John Britton, and all of the GitHub Staff for interviews, insights and ideas, Harold Solbrig, Craig Stancl, and Dick Hedger for guidance, and Rick Kiefer for his review.

@@

\input{interview}
\input{github_interview}
